Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

### 2023-04-24 Draft.Rmd
final

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
## Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

Air pollution has consequences for everyone, most especially for those with pre-existing conditions. Air pollution is measured in tiny particles or droplets in the air that are two and one-half microns or less in width called PM2.5. So being able to predict the average PM2.5 concentrations ($\mu$g/m$^3$) will allow us to keep people most vulnerable safe as we look for and implement changes to improve the air quality in the region. We created four models to predict the annual average of ambient air pollution in a location based on the variables in the dataset. We chose our predictors using the Random Forest algorithm. And for our four models, we chose Linear Regression, K-Nearest Neighbors, XG Boost, and Random Forest. Linear Regression models a linear relationship between the PM2.5 value in the atmosphere and the predictors. K-Nearest Neighbors returns predicted PM2.5 values based on their neighboring data points. XG Boost predicts PM2.5 values by combining the estimates of a set of simpler, weaker models. Random forest makes an averaged prediction from a collection of independent decision trees. We hypothesize that the RMSE performance of the XG Boost model will be the best, with a value of around 0.5.

```{r Library message=FALSE}
library(tidyverse)
library(tidymodels)
library(broom)
library(xgboost)
library(caret)
library(plotROC)
library(randomForest)
library(factoextra)
library(MASS)
library(ranger)
library(doParallel)

origin <- read_csv("https://github.com/rdpeng/stat322E_public/raw/main/data/pm25_data.csv.gz")
```

@@

```{r}
# Use Random Forest for variable selection
rfmodel = randomForest(value~., data=train[,2:50])
import=as.data.frame(randomForest::importance(rfmodel)) |> arrange(desc(IncNodePurity))
import=tibble::rownames_to_column(import, 'variables')

# take top 5 important variables
import5 = import[1:5,]
```

@@

wf1 <- workflow() %>% 
    add_recipe(rec1) %>% 
    add_model(model1)
# wf1

@@

# Check performance using cross-validation
folds2 <- vfold_cv(train_knn, v = 10)
# folds
res2 <- fit_resamples(wf2, resamples = folds2) # ERROR Assigned data `orig_rows` must be compatible with existing data.
res2 %>% 
    collect_metrics()
```

@@

# head(xgboost_grid)

@@

# Hyperparameter tuning
# Takes a long time to run, 5-8 min
xgboost_tuned <- tune::tune_grid(object = xgboost_wf, resamples = folds3,
  grid = xgboost_grid, metrics = yardstick::metric_set(rmse, rsq),
  control = tune::control_grid(verbose = TRUE))

# Find hyper parameter values which performed best at minimizing RMSE
xgboost_tuned %>% tune::show_best(metric = "rmse") %>%knitr::kable()
# Isolate the best performing hyperparameter values.
xgboost_best_params <- xgboost_tuned %>% tune::select_best("rmse")
xgboost_best_params

@@

# Tuning hyperparameters
all_cores = parallel::detectCores(logical=FALSE)
cl = makePSOCKcluster(all_cores)
doParallel::registerDoParallel(cl)

# Takes long time to load 2 min
tune_res <- wf4 |> tune_grid(resamples = folds4, grid = rf_grid, metrics=metric_set(rmse,rsq), control = tune::control_grid(verbose = TRUE))
# tune_res

# Find hyper parameter values which performed best at minimizing RMSE
tune_res %>% tune::show_best(metric = "rmse") %>%knitr::kable()
# Isolate the best performing hyperparameter values.
rf_best_params <- tune_res %>% tune::select_best("rmse")
rf_best_params

@@

```{r Q2}
# Visual for the most frequently close predictions by state
loc |> filter(diff<0.5) |> ggplot(aes(y=state)) + geom_bar()

# Visual for the most frequently furthest predictions by state
loc |> filter(diff>=0.5) |> ggplot(aes(y=state)) + geom_bar()
```

@@

head(xgboost_grid)

@@

# Find hyper parameter values which performed best at minimizing RMSE
xgboost_tuned %>% tune::show_best(metric = "rmse") %>%knitr::kable()
# Isolate the best performing hyperparameter values.
xgboost_best_params <- xgboost_tuned %>% tune::select_best("rmse")
xgboost_best_params

@@

**Include acknowledgments for any help received. If a group project, report the contribution of each member (i.e. who did what?).**
## Acknowledgements
**Aileen Li & Nyah Strickland:**
- Intro

>>>>

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
Air pollution has consequences for everyone, most especially for those with pre-existing conditions. Air pollution is measured in tiny particles or droplets in the air that are two and one-half microns or less in width called PM2.5. So being able to predict the average PM2.5 concentrations ($\mu$g/m$^3$) will allow us to keep people most vulnerable safe as we look for and implement changes to improve the air quality in the region. We created four models to predict the annual average of ambient air pollution in a location based on the variables in the dataset. We chose our predictors using the Random Forest algorithm. And for our four models, we chose Linear Regression, K-Nearest Neighbors, XG Boost, and Random Forest. Linear Regression models a linear relationship between the PM2.5 value in the atmosphere and the predictors. K-Nearest Neighbors returns predicted PM2.5 values based on their neighboring data points. XG Boost predicts PM2.5 values by combining the estimates of a set of simpler, weaker models. Random forest makes an averaged prediction from a collection of independent decision trees. We hypothesize that the RMSE performance of the XG Boost model will be the best, with a value of around 0.5.

```{r Library}
library(tidyverse);library(tidymodels);library(broom);library(xgboost);library(caret);library(plotROC)
library(randomForest);library(factoextra);library(MASS);library(ranger);library(doParallel)
origin <- read_csv("https://github.com/rdpeng/stat322E_public/raw/main/data/pm25_data.csv.gz")
```

@@

```{r}
# Use Random Forest for variable selection
rfmodel = randomForest(value~., data=train[,2:50])
import=as.data.frame(randomForest::importance(rfmodel)) |> arrange(desc(IncNodePurity))
import=tibble::rownames_to_column(import, 'variables')
# take top 5 important variables
import5 = import[1:5,]
```

@@

wf1 <- workflow() %>% 
    add_recipe(rec1) %>% 
    add_model(model1)

@@

# Check performance using cross-validation
folds2 <- vfold_cv(train_knn, v = 10)
# folds
res2 <- fit_resamples(wf2, resamples = folds2)
res2 %>% collect_metrics()
```

@@

@@

# Hyperparameter tuning, Takes a long time to run, 5-8 min
xgboost_tuned <- tune::tune_grid(object = xgboost_wf, resamples = folds3,
  grid = xgboost_grid, metrics = yardstick::metric_set(rmse, rsq),
  control = tune::control_grid(verbose = TRUE))

# Find hyper parameter values which performed best at minimizing RMSE
xgboost_tuned %>% tune::show_best(metric = "rmse")
# Isolate the best performing hyperparameter values.
xgboost_best_params <- xgboost_tuned %>% tune::select_best("rmse")
xgboost_best_params

@@

# Tuning hyperparameters
all_cores = parallel::detectCores(logical=FALSE)
cl = makePSOCKcluster(all_cores);doParallel::registerDoParallel(cl)

# Takes long time to load 2 min
tune_res <- wf4 |> tune_grid(resamples = folds4, 
                             grid = rf_grid, 
                             metrics=metric_set(rmse,rsq), 
                             control = tune::control_grid(verbose = TRUE))

# Find hyper parameter values which performed best at minimizing RMSE
tune_res %>% tune::show_best(metric = "rmse")
# Isolate the best performing hyperparameter values.
rf_best_params <- tune_res %>% tune::select_best("rmse")
rf_best_params

@@

```{r Q2}
# Visual for the most frequently close predictions by state
loc |>filter(diff<0.5) |> ggplot(aes(y=state)) + 
  geom_bar() + 
  labs(title='Most Frequent Close Predictions by State')

# Visual for the most frequently furthest predictions by state
loc |> filter(diff>=0.5) |> ggplot(aes(y=state)) + 
  geom_bar() + labs(title='Most Frequent Far Predictions by State')
```

@@

# Grid specification
xgboost_params <- dials::parameters(min_n(), tree_depth(), learn_rate(), loss_reduction())
xgboost_grid <- dials::grid_max_entropy(xgboost_params, size = 60)

@@

# Find hyper parameter values which performed best at minimizing RMSE
xgboost_tuned %>% tune::show_best(metric = "rmse")
# Isolate the best performing hyperparameter values.
xgboost_best_params <- xgboost_tuned %>% tune::select_best("rmse")
xgboost_best_params

@@

**Include acknowledgments for any help received. If a group project, report the contribution of each member (i.e. who did what?).**
## Acknowledgements
**Aileen Li & Nyah Strickland:**

- Intro
